---
title: "Visualization for Airplanes Crashes"
author: "Daniel Villalobos"
date: "2023-06-23"
output:
  ioslides_presentation:
    widescreen: yes
    css: styles.css
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
library(ggplot2)
library(lubridate)
library(dplyr)
library(tidyr)
library(gridExtra)
library(htmltools)
library(tmaptools)
library(wordcloud)
library(tm)
library(SnowballC)
```


```{r, include=FALSE}
setwd("/Users/danielvillalobostorrejon/Documents/Uni/VD")
filepath = "/Users/danielvillalobostorrejon/Documents/Uni/VD/PRA/Airplane_Crashes_and_Fatalities_Since_1908.csv"
df <- read.csv(filepath, sep=",")

filepath = "/Users/danielvillalobostorrejon/Documents/Uni/VD/PRA/region_data.csv"
region_data <- read.csv2(filepath)
```


```{r, include=FALSE}
df$Date <- as.Date(df$Date,  format = "%m/%d/%Y")
```

```{r, include=FALSE}
# Creating interesting variables
df$month <- month(df$Date, label = TRUE, abbr = TRUE)
df$day_of_week <- wday(df$Date, label = TRUE, abbr = TRUE)
df$hour <- hour(hm(df$Time))
df$year <- format(df$Date, "%Y")
df$is_militar <- ifelse(grepl("Military", df$Operator), "Militar", "Passenger")
df$country <- trimws(sub(".*,\\s*", "", df$Location))

# Filter bad values
df <- df %>% 
  filter(as.numeric(hour) <= 24)

x_labels <- seq(1910, 2009, 10)

```

```{r, include=FALSE}
# Create a vector of regions
regions <- df$country

# Initialize an empty dataframe to store the geospatial data
geocoded_data <- data.frame(region = character(),
                            lon = numeric(),
                            lat = numeric(),
                            stringsAsFactors = FALSE)

# Loop through each region and geocode it
# for (region in regions) {
#   # Geocode the region
#   tryCatch({
#     result <- geocode_OSM(region)
#     longitude <- result$coords[1]
#     latitude <- result$coords[2]
#     
#     # Check if the geocoding was successful
#     if (!is.na(longitude) && !is.na(latitude)) {
#       # Add the geocoded data to the dataframe
#       geocoded_data <- rbind(geocoded_data, data.frame(region = region,
#                                                        lon = longitude,
#                                                        lat = latitude,
#                                                        stringsAsFactors = FALSE))
#     } else {
#       # Set the coordinates of the United States as default
#       longitude <- -95.7129
#       latitude <- 37.0902
#       
#       # Add the geocoded data to the dataframe with default values
#       geocoded_data <- rbind(geocoded_data, data.frame(region = region,
#                                                        lon = longitude,
#                                                        lat = latitude,
#                                                        stringsAsFactors = FALSE))
#     }
#   }, error = function(e) {
#     # Handle the error if geocoding throws an error
#     cat("Geocoding failed for:", region, ". Error:", conditionMessage(e), "\n")
#   })
# }

# write.csv2(geocoded_data, "region_data.csv", row.names = FALSE)
```


## Total Crashes by Year

```{r, include=FALSE}
df_aggregated <- df %>%
  mutate(year = format(Date, "%Y")) %>%
  group_by(year) %>%
  summarise(total_crashes = n())

df_aggregated$row_index <- 1:nrow(df_aggregated)
```

- Tendencia creciente a tener más accientes 
- También ha aumentado el uso de los aviones para realizar viajes

```{r, warning=FALSE}
ggplot(df_aggregated, aes(x = year, y = total_crashes)) +
  geom_point() +
  geom_segment(aes(x = year, xend = lead(year), y = total_crashes, yend = lead(total_crashes)),
               color = "lightblue", size = 0.3) +
  labs(x = "Year", y = "Total Crashes") +
  ggtitle("Total Crashes by Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_x_discrete(breaks = as.character(x_labels))
```


## Total Crashes Periodically

* No se observa ningún més con una tendencia mayor a los demás
* Tampoco se observa una tendencia mayor en ninguno de los días de la semana

```{r, include=FALSE}
# Count crashes by month
df_aggregated <- df %>%
  group_by(month) %>%
  summarise(total_crashes = n())
```


```{r, warning=FALSE}
# Plot crashes by month as a bar plot
p1 <- ggplot(df_aggregated, aes(x = month, y = total_crashes)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(x = "Month", y = "Total Crashes") +
  ggtitle("Total Crashes by Month") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, include=FALSE}
# Count crashes by day of week
df_aggregated <- df %>%
  group_by(day_of_week) %>%
  summarise(total_crashes = n())

# Reorder the days of the week
df_aggregated$day_of_week <- factor(df_aggregated$day_of_week, levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))
```


```{r, warning=FALSE}
# Plot crashes by day of week as a bar plot
p2 <- ggplot(df_aggregated, aes(x = day_of_week, y = total_crashes)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(x = "Day of Week", y = "Total Crashes") +
  ggtitle("Total Crashes by Day of Week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
grid_arrange_shared <- grid.arrange(p1, p2, ncol = 2)
```



```{r, include=FALSE}
df_aggregated <- df %>%
  group_by(hour) %>%
  summarise(total_crashes = n())
```


<!-- ## Total Crashes Periodically -->

```{r, warning=FALSE, include=FALSE}
ggplot(df_aggregated, aes(x = hour, y = total_crashes)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(x = "Hour", y = "Total Crashes") +
  ggtitle("Total Crashes by Hour") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Total Crashes Per Operator

* Se observa como en el global de accidentes de avión la mayoria son de operadoras comerciales
* Existe un período donde los accidentes militares son más comunes que no los accidentes comerciales

```{r, include=FALSE}
category_props <- df %>%
  group_by(is_militar) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count),
         label = paste0(round(prop*100,0), "% (", count, ")"))
```

```{r}
p1 <- ggplot(category_props, aes(x = "", y = prop, fill = is_militar)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(fill = "is_militar") +
  ggtitle("Distribution of Flights") +
  theme_minimal() +
  theme(legend.position = "right") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5))+ 
  guides(fill = "none")
```

```{r, include=FALSE}
df_summary <- df %>%
  group_by(year, is_militar) %>%
  summarise(total_accidents = n())
```

```{r}
p2 <- ggplot(df_summary, aes(x = year, y = total_accidents, color = is_militar, group = is_militar)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Total Accidents", color = "Category") +
  ggtitle("Yearly Accidents per Category") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_x_discrete(breaks = as.character(x_labels))
```
```{r}
grid.arrange(p2, p1, ncol=2, widths = c(2, 1))
```


## World Crashes

```{r, include=FALSE}
crashes_per_country <- aggregate(. ~ country, df, length)
names(crashes_per_country)[names(crashes_per_country) == "country"] <- "region"
# Merge crash data with region data
merged_data <- merge(crashes_per_country, region_data, by = "region", all.x = TRUE)
# Load world map data
world_map <- map_data("world")
```

```{r, warning=FALSE}
# Plot the world map with crashes data
p1 <- ggplot() +
  geom_map(data = world_map, map = world_map,
           aes(x = long, y = lat, map_id = region),
           fill = "lightgray", color = "gray", size = 0.25) +
  geom_map(data = merged_data, map = world_map,
           aes(fill = Fatalities, map_id = region),
           color = "gray", size = 0.25) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Crashes") +
  labs(title = "Countries with Highest Number of Crashes") +
  theme_void()
```

* En este mapamundi de calor podemos observar aquellas regiones con más accidentes
* Rusia y Canada son las regiones donde historicamente se han producido más accidentes

```{r}
p1
```

## Top 10 regions

```{r, include=FALSE}
df_summary <- df %>%
  group_by(country) %>%
  summarise(total_crashes = n())

# Filter and select the top 10 countries
top_10 <- df_summary %>%
  top_n(10, total_crashes) %>%
  arrange(desc(total_crashes))

# Calculate the total number of crashes
total_crashes <- sum(top_10$total_crashes)

# Calculate the percentage of crashes for the top 10 countries
```


```{r}
# Plot the bubble graph
p1 <- ggplot(top_10, aes(x = reorder(country, total_crashes), y = total_crashes, size = total_crashes)) +
  geom_point(color = "blue", alpha = 0.7) +
  labs(x = "Country", y = "Number of Crashes", title = "Top 10 Regions with the Highest Number of Crashes") +
  theme_minimal()
```


```{r, include=FALSE}
# Create a separate big number graph
total_percentage <- total_crashes / sum(df_summary$total_crashes) * 100

big_number <- data.frame(label = paste0(round(total_percentage, 1), "%"))
```


```{r}
p2 <- ggplot(big_number, aes(x = 1, y = 1, label = label)) +
  geom_text(size = 50, color = "blue") +
  xlim(0.5, 1.5) +
  ylim(0.5, 1.5) +
  theme_void()
```

Este top 10 de regiones representa un **`r big_number`** de las colisiones totales de aviones

```{r}
p1
```



## Total Fatalities by Airplane Crashes per Year

```{r, include=FALSE}
df_aggregated <- df %>%
  group_by(year) %>%
  summarise(total_fatalities = sum(Fatalities),
            avg_fatalities = mean(Fatalities))

df_aggregated$row_index <- 1:nrow(df_aggregated)
```

* Aumento de las muertes por accidentes de avión


```{r, echo = FALSE, warning=FALSE}
# Plotting
ggplot(df_aggregated, aes(x = year, y = total_fatalities, fill = total_fatalities)) +
  geom_point() +
  geom_segment(aes(x = year, xend = lead(year), y = total_fatalities, yend = lead(total_fatalities)),
               color = "lightblue", size = 0.3) +
  labs(x = "Year", y = "Total Fatalities") +
  ggtitle("Total Fatalities by Airplane Crashes per Year") + 
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  guides(fill = "none")+
  scale_x_discrete(breaks = as.character(x_labels))
```

## Average Fatalities by Airplane Crashes by Year

* Pero si comparamos la media de muertes por accidente vemos como la tendencia se mantiene estable al largo del periodo

```{r, warning=FALSE}
ggplot(df_aggregated, aes(x = year, y = avg_fatalities, fill = avg_fatalities)) +
  # geom_col() +
  geom_point() +
  geom_segment(aes(x = year, xend = lead(year), y = avg_fatalities, yend = lead(avg_fatalities)),
               color = "lightblue", size = 0.3) +
  labs(x = "Year", y = "Total Fatalities") +
  ggtitle("Average Fatalities by Airplane Crashes per Year") + 
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  guides(fill = "none")+
  scale_x_discrete(breaks = as.character(x_labels))
```


## Evolution of Survivors by Year

* Además viendo como cada vez los aviones son más grandes y se usan más, es más probable que muera más gente

```{r}
df_summary <- df %>%
  group_by(year) %>%
  summarise(total_fatalities = sum(Fatalities),
            total_aboard = sum(Aboard))


df_summary <- df_summary %>%
  mutate(total_facilities = total_aboard - total_fatalities)

# Define the dodge width for positioning the bars
dodge_width <- 0.8

# Plot the grouped bar plot with legend
p1 <- ggplot(df_summary, aes(x = year)) +
  geom_bar(aes(y = total_aboard, fill = "Aboard"), stat = "identity", alpha = 0.5, position = position_dodge(width = dodge_width), width = dodge_width) +
  geom_bar(aes(y = total_facilities, fill = "Fatalities"), stat = "identity", alpha = 0.5, position = position_dodge(width = dodge_width), width = dodge_width) +
  scale_fill_manual(values = c("Aboard" = "green", "Fatalities" = "darkgreen"), labels = c("Aboard", "Fatalities")) +
  labs(x = "Year", y = "Count", fill = NULL) +
  ggtitle("Total Fatalities, and Aboard by Year") +
  theme_minimal()+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_x_discrete(breaks = as.character(x_labels))
```


```{r}
df_summary <- df %>%
  group_by(year) %>%
  summarise(total_fatalities = sum(Fatalities),
            total_aboard = sum(Aboard))

# Calculate the number of survivors
df_summary <- df_summary %>%
  mutate(total_survivors = (total_aboard - total_fatalities)/total_aboard*100)

# Plot the scatter plot
p2 <- ggplot(df_summary, aes(x = year, y = total_survivors, size = total_survivors, color = total_fatalities)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red") +
  labs(x = "Year", y = "Survivors", title = "Evolution of Survivors (Non-Military)") +
  theme_minimal() +
  guides(color = "none" , size = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_x_discrete(breaks = as.character(x_labels))
```

```{r}
p1
```

## Evolution of Survivors by Year

* Pero parece que la tasa de supervivencia es mayor actualmente que en el pasado

```{r}
p2
```


## Worldcloud Variable Summary


```{r, warning=FALSE}
# Combine all the summary text into a single character vector
text <- paste(df$Summary, collapse = " ")

# Create a corpus from the text
corpus <- Corpus(VectorSource(text))

# Preprocess the corpus
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removeWords, stopwords("english"))
corpus <- tm_map(corpus, stripWhitespace)

# Create a term document matrix
tdm <- TermDocumentMatrix(corpus)

# Convert the term document matrix to a matrix
matrix <- as.matrix(tdm)

# Convert the matrix to a data frame
word_freq <- data.frame(Terms = rownames(matrix), Frequency = matrix[, 1])
pal = brewer.pal(9,"Blues")
```

```{r, warning=FALSE}
wordcloud(word_freq$Terms, word_freq$Frequency, scale = c(5, 0.5), random.order = FALSE, colors = pal)
```

## 

<h2 style="text-align: center; 
           font-size: 80px; 
           align-items: center; 
           display: flex; 
           height: 60vh ">
           Gracias por su atención! 
</h2>

